Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2016-05-18 manuscript-final.tex
Rename figures with numbers
To not confuse the bad submission system.

Fixes #12
Had to remove figures and tables for submission

\documentclass[manuscript,endfloat]{geophysics}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

Satellite missions dedicated to measuring the Earth's gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
An analytical solution exists when the computation point is along
the polar axis and the tesseroid is extended into a spherical cap
\citep{Lafehr1991, Mikuska2006, Grombein2013}.
For more general cases,
the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists in approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that it can be
controlled by the number of point masses used.
The larger the number of point masses,
the better the accuracy of GLQ integration.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time.
This is a common theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

@@

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
        After \citet{uieda2015}.
    }
    \label{fig:tesseroid}
\end{figure}

A tesseroid is a mass element defined in geocentric spherical
coordinates
(Figure~\ref{fig:tesseroid}).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~\ref{fig:tesseroid}).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational acceleration,
and Marussi tensor, respectively,

@@

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two ($2 \times 2 \times 2 = 8$ point masses).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an ad hoc criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30
($30 \times 30 \times 30 = 27,000$ point masses).
The computed $g_{xy}$ component more closely resembles
the expect results for a single tesseroid \citep{Asgharzadeh2007}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point masses instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30.
        In (c) the horizontal locations of the point masses were not shown.
        Notice that the results shown in (c) are similar to that expected
        for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

Our definition of the dimensions of the tesseroid
(the ``side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

\noindent
$L_\lambda$ and $L_\phi$ are arc-distances measured along the top surface of
the tesseroid (Figure~\ref{fig:division}a).
Specifically, $L_\lambda$ is measured long the middle latitude of the
tesseroid ($\phi_t$).

@@

Figure~\ref{fig:division} shows examples of
the resulting tesseroid models after adaptive discretization.
Figure~\ref{fig:division}a shows
the initial tesseroid and computation point P.
Figures~\ref{fig:division}b-d are
the result of adaptive discretization using
different values of the distance-size ratio $D$,
respectively,
$D=1$, $D=2$, and $D=6$.
The number of tesseroids in the resulting discretization is, respectively,
4, 38, and 936.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

@@

The key controlling point of the adaptive discretization algorithm
is the distance-size ratio $D$ (equation~\ref{eq:condition}).
The specific value chosen for $D$ determines how many divisions will be made
(Figure~\ref{fig:division}).
Thus, $D$ indirectly controls both the accuracy of the integration
and the computation time.
In this section, we investigate the relationship between
the distance-size ratio and the integration error.
We perform the analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~\ref{tbl:experiment}).
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.

We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the error analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

Figure~\ref{fig:dist-size-curves} shows
the maximum difference between the shell and tesseroid fields
as a function of $D$ for the four experiments.
The differences are given as a percentage of the shell value.
We established a maximum tolerated error of $0.1\%$, represented by the
horizontal solid lines in Figure~\ref{fig:dist-size-curves}.
Only results for the gravitational potential, $g_z$, and $g_{zz}$ are shown.
The results for the other diagonal components of the gravity gradient tensor
are similar to $g_{zz}$.
Figures for these components can be found in the supplementary material
(see section "Reproducing the analysis and results").

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

\input{table1.tex}

\begin{figure}
    \centering
    \includegraphics{figs/distance-size-curves}
    \caption{
        The maximum difference between the computed tesseroid and shell effects
        as a function of the distance-size ratio $D$
        for (a) the gravitational potential, (b) $g_z$, and (c) $g_{zz}$.
        The difference is given as a percentage of the shell effect.
        Curves correspond to the different tesseroid models and computation
        grids shown in Table~\ref{tbl:experiment}.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:dist-size-curves}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figs/gzz-with-height}
    \caption{
        Difference between the computed $g_{zz}$ for the spherical shell and
        the tesseroid model at different heights. Curves show the maximum
        difference as a percentage of the shell value.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:gzz-with-height}
\end{figure}

@@

>>>>

% Final version submitted to the journal system. Had to remove all
% cross-references to figures and tables and include only the captions.
% Bibliography is included in this file, instead of the .bib file.

\documentclass[manuscript]{geophysics}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

Satellite missions dedicated to measuring the Earth's gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~1)
instead of rectangular prisms.
An analytical solution exists when the computation point is along
the polar axis and the tesseroid is extended into a spherical cap
\citep{Lafehr1991, Mikuska2006, Grombein2013}.
For more general cases,
the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists in approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that it can be
controlled by the number of point masses used.
The larger the number of point masses,
the better the accuracy of GLQ integration.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time.
This is a common theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

@@

A tesseroid is a mass element defined in geocentric spherical
coordinates
(Figure~1).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~1).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational acceleration,
and Marussi tensor, respectively,

@@

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~2
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~2a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two ($2 \times 2 \times 2 = 8$ point masses).
Figure~2b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an ad hoc criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~2c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30
($30 \times 30 \times 30 = 27,000$ point masses).
The computed $g_{xy}$ component more closely resembles
the expect results for a single tesseroid \citep{Asgharzadeh2007}.

@@

Our definition of the dimensions of the tesseroid
(the ``side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~3a)

@@

\noindent
$L_\lambda$ and $L_\phi$ are arc-distances measured along the top surface of
the tesseroid (Figure~3a).
Specifically, $L_\lambda$ is measured long the middle latitude of the
tesseroid ($\phi_t$).

@@

Figure~3 shows examples of
the resulting tesseroid models after adaptive discretization.
Figure~3a shows
the initial tesseroid and computation point P.
Figures~3b-d are
the result of adaptive discretization using
different values of the distance-size ratio $D$,
respectively,
$D=1$, $D=2$, and $D=6$.
The number of tesseroids in the resulting discretization is, respectively,
4, 38, and 936.

@@

The key controlling point of the adaptive discretization algorithm
is the distance-size ratio $D$ (equation~\ref{eq:condition}).
The specific value chosen for $D$ determines how many divisions will be made
(Figure~3).
Thus, $D$ indirectly controls both the accuracy of the integration
and the computation time.
In this section, we investigate the relationship between
the distance-size ratio and the integration error.
We perform the analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

@@

Figure~2 shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(see Table~1).
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.

We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.
Furthermore, we repeated the error analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.
Table~1 describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

Figure~4 shows
the maximum difference between the shell and tesseroid fields
as a function of $D$ for the four experiments.
The differences are given as a percentage of the shell value.
We established a maximum tolerated error of $0.1\%$, represented by the
horizontal solid lines in Figure~4.
Only results for the gravitational potential, $g_z$, and $g_{zz}$ are shown.
The results for the other diagonal components of the gravity gradient tensor
are similar to $g_{zz}$.
Figures for these components can be found in the supplementary material
(see section "Reproducing the analysis and results").

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~5 shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

@@

\newpage
\section{Figure captions}

\begin{enumerate}
    \item
    View of a tesseroid, the integration point $Q$ inside the tesseroid, a
    geocentric coordinate system $(X, Y, Z)$, the computation $P$ and it's
    local coordinate system $(x, y, z)$.  $r$, $\phi$, $\lambda$ are the
    radius, latitude, and longitude, respectively, of point $P$, and $\ell$ is
    the Cartesian distance between $P$ and $Q$.  After \citet{uieda2015}.

    \item
    Example of the effect of varying the computation height and the number of
    point masses in the Gauss-Legendre Quadrature.  Black circles represent the
    horizontal location of the point masses.  a) $g_{xy}$ calculated at $400\
    km$ height using GLQ order 2 ($2 \times 2 \times 2 = 8$ point masses).  b)
    At $150\ km$ height and GLQ order 2, the result resembles that of four
    point masses instead of a single tesseroid.  This effect was shown by
    \citet{Ku1977}.  c) At $150\ km$ but with a higher GLQ order of 30.  In (c)
    the horizontal locations of the point masses were not shown.  Notice that
    the results shown in (c) are similar to that expected for a single mass
    source.

    \item
    Adaptive discretization of the tesseroid shown in (a) for a computation
    point P using the distance-size ratio $D$ equal to (b) 1, (c) 2, and
    (d) 6.  $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the
    tesseroid.  Note that increasing $D$ results in a fine division of the
    tesseroid close the computation point and a coarser division further
    away.

    \item
    The maximum difference between the computed tesseroid and shell effects
    as a function of the distance-size ratio $D$ for (a) the gravitational
    potential, (b) $g_z$, and (c) $g_{zz}$.  The difference is given as a
    percentage of the shell effect.  Curves correspond to the different
    tesseroid models and computation grids shown in Table~1.  The
    horizontal solid black line marks the established error threshold of
    $0.1\%$.  A value of $D=0$ means that no divisions are made.

    \item
    Difference between the computed $g_{zz}$ for the spherical shell and
    the tesseroid model at different heights. Curves show the maximum
    difference as a percentage of the shell value.  The horizontal solid
    black line marks the established error threshold of $0.1\%$.  A value
    of $D=0$ means that no divisions are made.

\end{enumerate}

\newpage
\section{Table captions}

\begin{enumerate}
    \item
    Parameters of the numerical experiments to quantify the accuracy of the
    numerical integration.  All grids had $10\times10$ regularly spaced
    computation points at a constant height.  Tesseroids used to discretize the
    spherical shell had 1 km thickness and the horizontal dimensions shown in
    the table.
\end{enumerate}
